---
title: "Part 1 Data Visualisation using R: giving a CRAP with ggplot"
output: html_notebook
author: C E Page, F Robinson
---


What do we need to consider when creating visualisations of our data?

1. Begin by defining the purpose of your visual(s), and its intended audience.

2. Choose appropriate form(s) for the intended purpose and function. **A LITTLE BIT OF THIS**

3. Consider how best to visualise the data using design principles (CRAP). **THIS WORKSHOP**


***
# Part 1: giving a CRAP with ggplot 
***


Goals of Part 1:


1. __ggplot__ - how building a `ggplot` works - step by step.


2. __CRAP & ggplot?__ - core graphical principles (CRAP - Contrast, Repition, Alignment and Proximity) that when applied to data in R can help visualise science in a meaningful (and beautiful) way!


3. __what plot?__ - choosing what plot is also key to great data visualisation! This deserves a whole tutorial in itself, so today we just point to some of the options you have in R, and resources you can consult for further direction.


```{r eval=T, echo=T, message = FALSE}
# Load libraries 
# if you need to install these run the code install.packages("package_name")

library(tidyverse)
library(ggthemes)
library(ggrepel) # for annotations 
library(viridis) # for color blind friendly colours 
library(RColorBrewer) 
library(reshape2)

```


***
### 1. __ggplot__ 
***

**Fiona please feel free to take or leave what you want from the below - I shan't be offended :) )**

`ggplot2` is a powerful R package developed by Hadley Wickham (REF). It is part of the tidyverse and is arguably most effective for creating publication quality graphics. There are many extension packages for ggplot which can make it easy to reproduce specialised types of graphs (e.g. survival plots, geographic maps).


The gg in `ggplot2` stands for the Grammar of Graphics - this is a graphics concept which describes plots by using a specified "grammar".  


A plot can be divided into several fundamental parts:

1. __data__ a tidy data frame.
2. __aes__ visual properties that get mapped onto values in a data set (colour, size, shape etc.). This inclues specifying the variables to be plotted (x,y).
3. __geom__ corresponds to the type of plot you are trying to make (histogram, box plot, line plot, density plot, dot plot...).
4. __stat__ a stat layer applies some kind of statistical transformation to the data. For example `stat_smooth(method = "lm")` displays a linear regression line with confidence intervals.
5. __theme__ a theme is a set of visual parameters that controls the background, borders, text, legend position etc. You can use predefined themes or create your own (see below!).

There are two main funtions for creating plots in the ggplot2 package: a `qplot()` and `ggplot()` function: 

+ `qplot()` quick plot function easy to use for simple plots
+ `ggplot()` more flexible function that allows you to build a plot piece by piece

Plots can be saved as a variable using <- assign, and printed at any time by calling the variable. 

After creating a plot use `ggsave` to save a plot in the current working directory. 



##### Coding a ggplot



The key to understanding `ggplot2` is thinking about a figure in layers built using these multiple componants (the core being the data, aesthetics and geom) **(NOTE - I think it would be nice to do a live code example here! We could take participants through building a really simple ggplot with each of these aspects of the plot. i.e. first you specify the data (tidy data, what is tidy data for ggplot? LONG format), and then set up the aesthetics...etc.). It would be great if we could also show how to edit themes using theme() as an introduction to specifying a theme. Maybe it would also be worth showing how to save ggplots as pdfs, jpgs etc.**

ggplot follows this template: **(NOTE - I have currently specificed x and y as aes within each geom. I figure this makes understanding how to add multiple plot types into one easier to understand for part 2 rain cloud plot.)**


```{r eval=F, echo=T}

ggplot(data = DATA) + 
  geom_function(mapping = aes(AESTHETIC MAPPINGS))

```


***
### 2. __CRAP & ggplot__ 
***

We can look to use some simple core graphical principles (CRAP - Contrast, Repition, Alignment and Proximity) that when applied to can help visualise data in a beautiful (and meaningful) way! The CRAP principles of design were developed by Robin Williams (REF) and refur to the four basic principles of design that xxxx.


**(CP to provide more background here, in addition to links to other useful resources - also add graphic detailing what each of the main words mean)**


Some of these principles we can look to directly apply during graphics creation in R, whereas others can be applied during figure creation, or even during production of graphical abstracts. 


#### C - Contrast 


If two items are not exactly the same, make them different! Applying contrast can help attract attention to important elements, and groups of elements by making them stand out. 


We can look to do this in plots by emphasising group difference with colour and size, in addition to using different font families, and font sizes for plot annotation. We will learn how to specify fonts and font sizes in ggplot theme() in section R - Repetition below. There are many fonts, and font weights (i.e. **bold**, *italic*, light) you can specify in R. 

Below is code that creates a plot of some example fonts and font weights you can specify in R:

```{r}

```

```{r}

```


Colour is also a powerful tool and can be used to explain levels of data (e.g. through application of a gradient) in addition to simple, and more complex hierachical grouping structures. 

Some great online tools for choose palettes based on colour theory: 


+ adobe colour 


When we think about including colour in our graphics we need to make sure we think about useability and accessability; where we make sure that the palette (range of colours) we choose are not misleading, and are differentiatable for those with varying colourblindness.


To help deal with problems associated with useability and accessability there are a number of really useful packages developed in R (in addition to a colour blind checker on adobe colour!). These include: 


  + Greys

  Sometimes absence of colour can be more powerful in communicating ideas. R has a number of built in grey palettes. 
  
  Usage in ggplot: 
      
        + scale_fill_grey() for box plot, bar plot, violin plot, dot plot etc.
        
        + scale_color_grey() for lines and points
        

+ RColorBrewer palettes
  
  The RColorBrewer package creates great palettes - that can also be filtered to be colourblind friendly.
  
  The package contains 3 sets of color palette: sequential, qualitative and diverging. Understanding these groupings of    colour can help in choosing how to include colour within your plots.
  
  1. Sequential - ordered data that progress from low to high.
  
  2. Qualitative - best suited for nominal or categorical data where colour does not imply magnitude between groups.
  
  3. Diverging - these put equal emphasis on mid-range and extreme values at both ends of the data range.
  
  Below are a few useful commands in RColorBrewer:
  
 
```{r}
# Display all colour palettes 
display.brewer.all()

# Display only colourblind friendly palettes
display.brewer.all(n = NULL, type = "all", select = NULL, # type = "div","qual","seq"
                   colorblindFriendly = TRUE)

#  Return the hexadecimal color specification 
# brewer.pal(n, name) # where n is the number of colours you want from the palette - minimum is 3!

# Display a single RColorBrewer palette 
# by specifying its name
# display.brewer.pal(n, name)

```

  Usage in ggplot
  
  
      + scale_fill_brewer() for box plot, bar plot, violin plot, dot plot etc.
      
      + scale_color_brewer() for lines and points


We can also change the size of the elements in a graph - which in itself can present data values.



#### R - Repetition 


Repeat things in your design, throughout your design. This helps creats a consistent look and feel, where recurring elements can help organise content and establish a sense of unity. 

How? There are a number of ways we can bring repitiion into our designs in R, namely through the use of themes.  


#### A - Alignment 


Nothing should be placed randomly on a page. Make sure that every item has some visual connection with something else.  

How? In R we can look to align graphics we create using cowplot() (ref to other tutorial/resource) and patchwork().


#### P - Proximity

Group graphics together that are related. An important benefit here is that grouping graphics can organise information, and clarify concepts. This concept can be applied when using patchwork() to create groups of figures. 

Tip: Try and group plots together from the same hypothesis group in your study, or plots that show results from the same method. 




